Resultants and subresultants of $p$-adic polynomials

### 2015-06-25 resultants/resultants.tex
Talk at ISSAC: preliminary version of slides

\usepackage{amsmath,amssymb,amsthm}
%\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{multirow}
\renewcommand*{\theHsection}{\thesection}

@@

\newcommand{\ring}{\mathfrak A}
\newcommand{\fracring}{\mathfrak K}
\renewcommand{\prec}{\text{\rm prec}}
\renewcommand{\val}{\text{\rm val}}
\newcommand{\id}{\text{\rm id}}
\newcommand{\Res}{\text{\rm Res}}
\newcommand{\lc}{\text{\rm lc}}
\newcommand{\Card}{\text{\rm Card}\:}

@@

\vspace{-0.3cm}
\end{figure}
%
Figure \ref{fig:precision} illustrates the instability of extended Euclide's 
algorithm when it is called on random inputs which are monic $2$-adic 
polynomials of fixed degree. \todo{(Donner l'algorithme.)} Looking at the last line, we see that 
extended Euclide's algorithm outputs the B\'ezout coefficients of two 
monic $2$-adic polynomials of degree $100$ with an average loss of 
$161.1$ significant digits by coefficient whereas a stable 
algorithm should only loose $3.2$ digits on average. This 
``theoretical'' loss is computed as the double of the valuation of the 
resultant. Indeed Cramer-like formulae imply that B\'ezout coefficients 
can be computed by performing a unique division by the resultant, 
inducing then only the aforementioned loss of precision. Examining the 
table a bit more, we furthermore observe that the ``practical'' loss of 
precision due to Euclide's algorithm seems to grow linearly with respect 
to the degree of the input polynomials whereas the ``theoretical'' loss 
seems to be independant of it. In other words, the instability of 
Euclide's algorithm is becoming more and more critical when the degree 
of the input increases.

@@

A first quite important class of examples of discrete valuation rings 
are rings of formal power series over a field. They are equipped with 
the standard valuation defined as follows: $\val(\sum_{i \geq 0} a_i 
t^i)$ is the smallest integer $i$ with $a_i \geq 0$. The corresponding 
distance on $k[[t]]$ is complete. Indeed, denoting by $f[i]$ the term in 
$t^i$ in a series $f \in k[[t]]$, we observe that a sequence $(f_n)_{n 
\geq 0}$ is Cauchy if and only if the sequences $(f_n[i])_{n \geq 0}$ 
are all ultimately constant. A Cauchy sequence $(f_n)_{n \geq 0}$ 
therefore converges to $\sum_{i \geq 0} a_i t^i$ where $a_i$ is the 
limit of $f_n[i]$ when $n$ goes to $+\infty$.
The DVR $k[[t]]$ has a distinguished uniformizer, namely $t$. Its 
maximal ideal is then the principal ideal $(t)$ and its residue field
is canonically isomorphic to $k$. 
If one chooses $\pi = t$ and constant polynomials as representatives of 
classes modulo $t$, the expansion \eqref{eq:expandCDVR} is nothing but 
the standard writing of formal series.
The fraction field of $k[[t]]$ is the ring of Laurent series over $k$
and, once again, the expansion \eqref{eq:expandCDVF} corresponds to the
usual writing of Laurent series.

@@

\todo{Détailler cette partie}
% parler notamment de précision absolue et de précision relative

@@

We recall briefly basic definitions and results about resultants and 
subresultants. For a more complete exposition, we refer to \cite[\S 
3.3]{cohen} and \cite[\S 4.1]{winkler}.
Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
$d_A$ and $d_B$ be two integer greater than or equal to the degree of 
$A$ and $B$ respectively. We consider the Sylvester application:
$$\begin{array}{rcl}
\psi : \quad
\ring_{< d_B}[X] \times \ring_{< d_A}[X] & \to &
\ring_{< d_A+d_B}[X] \smallskip \\
(U,V) & \mapsto & AU + BV
\end{array}$$
The Sylvester matrix is the matrix of $\psi$ in the canonical ordered
basis, which are
$$\begin{array}{rrl}
& ((X^{d_B-1},0), \ldots, (X,0), (1,0), (0,X^{d_A-1}), \ldots, (0,1))
& \text{for the source} \smallskip \\
\text{and} & (X^{d_A+d_B-1}, \ldots, X, 1) &
\text{for the target.}
\end{array}$$
The \emph{resultant} of $A$ and $B$ (computed in degree $d_A, d_B$) is the 
determinant of the $\psi$; we denote it by $\Res^{d_A,d_B}(A,B)$. We 
observe that it vanishes if $d_A > \deg A$ or $d_B > \deg B$. In what 
follows, we will freely drop the exponent $d_A, d_B$ if $d_A$ and $d_B$ 
are the degrees of $A$ and $B$ respectively.
Using Cramer formulae, we can build polynomials $U^{d_A, d_B}(A,B) \in
\ring_{< d_B}[X]$ and $V^{d_A, d_B}(A,B) \in \ring_{< d_A}[X]$ whose 
coefficients are, up to a sing, maximal minors of the Sylvester matrix 
such that:
$$A \cdot U^{d_A, d_B}(A,B) + B \cdot V^{d_A, d_B}(A,B) = 
\Res^{d_A, d_B}(A,B).$$
There polynomials are called the \emph{cofactors} of $A$ and $B$
(computed in degree $d_A, d_B$).

@@

Since they are defined as determinants, subresultants behave well with 
respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
and $A$ and $B$ are polynomials over $\ring$ then
$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$
where $f(A)$ and $f(B)$ denotes the polynomials deduced from $A$
and $B$ respectively by applying $f$ coefficient-wise. This property
is sometimes referred to as the \emph{functoriality} of subresultants.

@@

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above and assume that $d_A = \deg A$. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
$$\left\{\begin{array}{ll}
S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\delta_i + 1}
s_{i-1}^{-1} \: c_i^{-\delta_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 0 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\delta_{i+1}} \cdot c_i^{1-\delta_{i+1}}
& \text{for } i \geq -1. 
\end{array}\right.$$
Here $n_i = \deg S_i$ \todo{define $\delta_i$ and change the notation} 
and $s_i$ is the leading coefficient of 
$S_i$ if $i \geq 0$ and $s_{-1} = 1$ by convention. These sequences 
are finite and the above recurrence applies until $S_i$ has reached 
the value $0$.

\begin{prop}
\label{prop:prem}
With the above notations, we have:
$$\begin{array}{rcll}
\Res_j(A,B) & = & S_i & \text{if } j = n_{i-1} - 1 \smallskip \\
& = & 0 & \text{if } n_i < j < n_{i-1} - 1 < j \\
& = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\delta_i-1} \cdot S_i & 
\text{if } j = n_i
\end{array}$$
for all $i$ such that $S_i$ is defined.
\end{prop}

\begin{rem}
The Proposition \ref{prop:prem} computes \emph{all} subresultants. We
note moreover that, in the common case where $n_{i-1} = n_i - 1$, the
two formulas giving $\Res_{n_i}(A,B)$ agree.

Mimicing ideas behind extended Euclide's algorithm, one can define the 
``extended subresultant pseudo-remainder sequence'' as well and compute
this way at the same time the cofactors.
\end{rem}

Important simplifications occur in the common case where all principal 
subresultants do not vanish. Under this additional assumption, one can 
prove that the degrees of the $S_i$'s decrease by one at each step, 
meaning that we $\deg S_i = d_B - i$ for all $i$. The sequence 
$(S_i)$ then stops at $i = d_B$. Moreover, the $\delta_i$'s and the 
$c_i$'s are now all ``trivial'' since we have $\delta_i = 1$ and $c_i
= s_i$ for all $i$. The recurrence formula then becomes:
$$S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
\cdot (S_{i-1} \,\%\, S_i) \quad \text{for } i \geq 1.$$
To conclude with, we remark that Proposition \ref{prop:prem} simply 
states that $R_j = S_{d_B-j}$ in this case.

@@

%The aim of this section is to study the loss of precision

% Expliquer que les pertes avec l'algorithme des pseudo-restes
% sont les mêmes que ceux avec l'algorithme d'Euclide

@@

In this subsection, we estimate the loss of precision if we compute the 
$R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
In what follows, we are going to use a \emph{flat precision model}: this 
means that a polynomial $P(X)$ is represented as:
$$P(X) = \sum_{i=1}^n a_i X^i + O(\pi^N)
\quad \text{with } a_i \in K \text{ and } N \in \Z.$$
In other words, we assume that the software we are using does not
carry a precision data on each coefficient but only a unique precision
data for the whole polynomial. Concretely this means that, after having
computing a polynomial, the software truncates the precision on each
coefficients to the smallest precision. 
One can argue that this assumption is too strong (compared to usual
implementations of $p$-adic numbers). Nevertheless, it defines a 
simplified framework where computations can be performed and experiments 
show that it rather well reflects the behaviour of the loss of precision 
in Euclide-like algorithms.

Let $V_j$ be the valuation of the principal $j$-th subresultant of $A,B$ 
and $W_j$ be the minimum of the valuations of the coefficients of $R_j$. 
We of course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
\todo{Change these notations.}

@@

\begin{proof}
By functoriality of subresultants, we may assume that $\ring = 
\Z[a_0, \ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$
are the two generic monic polynomials 
$A = X^d + \sum_{i=0} a_i X^i$ and
$B = X^d + \sum_{i=0} b_i X^i$.
All relations then follows from the extended subresultants 
pseudo-remainder sequence algorithm (noting in addition that all
principal subresultants are nonzero).
\end{proof}

@@

\begin{proof} 
Given a subset $J$ of $\{0, \ldots, d-1\}$, we denote by 
$\Omega_\ring(J)$ the subset of $\Omega_\ring$ consisting of those 
couples $(A,B)$ for which $r_j(A,B)$ does not vanish if and only if $j 
\in J$. We want to prove that $\Omega_\ring(J)$ has cardinality 
$q^{2d-\Card J} (q-1)^{\Card J}$. To do this, we introduce several 
additional notations. First, we fix $J$ and write $J = \{n_1, \ldots, 
n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$. We also set $n_{\ell+1} = 
0$ by convention. Then, given $n$ and $m$ two integers with $m < n$, let 
$V_{(m,n)}$ denote the set of polynomials of the form
$a_m X^m + a_{m+1} X^{m+1} \cdots + a_n X^n$
with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{(m,n)}$ has
cardinality $(q-1) q^{n-m}$. If $P$ is any polynomial of 
degree $n$ and $m < n$ is an integer, we further define $P[m{:}] \in 
V_{(m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
with coefficients $< m$. Third, given $(A,B)$ in $\Omega_\ring$, we 
denote by $(S_i(A,B))$ its subresultant pseudo-remainder sequence as 
defined in \S \ref{subsec:subres}. We note that, if $(A,B) \in 
\Omega_\ring(J)$, the sequence $(S_i(A,B))$ stops at $i = \ell$ and we 
have $\deg S_i = n_i$ for all $i$. We now claim that the mapping
$$\begin{array}{rcl} 
\Lambda_J : \,\, 
\Omega_\ring(J) & \to & 
V_{(n_1,n_2)} \times \cdots \times V_{(n_\ell,n_{\ell+1})} \smallskip \\
(A,B) & \mapsto & 
\big(S_i(A,B)[n_{i+1}{:}]\big)_{1 \leq i \leq \ell}
\end{array}$$ 
is injective. In order to establish the claim, we remark that the 
knowledge of $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$ (for some $i$) 
is enough to reconstruct the quotient of the Euclidean division of 
$S_i(A,B)$ by $S_{i-1}(A,B)$. Thus, one can recover $S_i(A,B)$ from 
$S_{i-2}(A,B)$, $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$. We deduce 
from this that $\Lambda_J(A,B)$ determines uniquely all $S_i(A,B)$'s 
and finally $A$ and $B$ themselves. This proves the claim.

Finally, we note that the claim implies that the cardinality of 
$\Omega_\ring(J)$ is at most $q^{2d-\ell} (q-1)^\ell$. Summing up these 
inequalities over all possible $J$, we get $\Card \Omega_\ring \leq 
q^{2d}$. This latest inequality being an equality, we need to have
$\Card \Omega_\ring(J) = q^{2d-\Card J} (q-1)^{\Card J}$ for all $J$.
\end{proof}

@@

We have seen in the previous sections that Euclide's algorithm (and 
their variants) are very unstable in practice. On the other hand, 
one can compute subresultants in a very stable way by evaluating
the corresponding minors of the Sylvester matrix. Doing so, we do
not loose any significant digit. Of course, the downside is the
quite bad efficiency.

@@

In this subsection, we briefly report on the results of \cite{padicprec}. In this 
paper, the authors draw the lines of a general framework to handle a 
sharp (often optimal) track of ultrametric precision. In order to do so, 
they proposed to attach a single precision datum --- which is a lattice 
in a huge vector space --- to the set of all variables, instead of 
attaching a precision datum to each individual variable as we do 
usually. This proposal is supported by the following lemma in ultrametric
analysis.

@@

>>>>

\usepackage{amsmath,amssymb,amsthm}
%\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{multirow}
\renewcommand*{\theHsection}{\thesection}

@@

\newcommand{\ring}{\mathfrak A}
\newcommand{\fracring}{\mathfrak K}
\renewcommand{\prec}{\text{\rm prec}}
\renewcommand{\val}{\text{\rm val}}
\newcommand{\id}{\text{\rm id}}
\renewcommand{\Res}{\text{\rm Res}}
\newcommand{\lc}{\text{\rm lc}}
\newcommand{\Card}{\text{\rm Card}\:}

@@

\vspace{-0.3cm}
\end{figure}
%
Figure \ref{fig:precision} illustrates the instability of the classical
extended Euclide's algorithm (\emph{cf} Algorithm \ref{algo:Euclide})
when it is called on random inputs which are monic $2$-adic 
polynomials of fixed degree. Looking at the last line, we see that 
extended Euclide's algorithm outputs the B\'ezout coefficients of two 
monic $2$-adic polynomials of degree $100$ with an average loss of 
$161.1$ significant digits by coefficient whereas a stable 
algorithm should only loose $3.2$ digits on average. This 
``theoretical'' loss is computed as the double of the valuation of the 
resultant. Indeed Cramer-like formulae imply that B\'ezout coefficients 
can be computed by performing a unique division by the resultant, 
inducing then only the aforementioned loss of precision. Examining the 
table a bit more, we furthermore observe that the ``practical'' loss of 
precision due to Euclide's algorithm seems to grow linearly with respect 
to the degree of the input polynomials whereas the ``theoretical'' loss 
seems to be independant of it. In other words, the instability of 
Euclide's algorithm is becoming more and more critical when the degree 
of the input increases.
%
\begin{algorithm}
\todo{Write it}

\caption{Extended Euclide's algorithm}
\label{algo:Euclide}
\end{algorithm}

@@

A first class of examples of discrete valuation rings 
are rings of formal power series over a field. They are equipped with 
the standard valuation defined as follows: $\val(\sum_{i \geq 0} a_i 
t^i)$ is the smallest integer $i$ with $a_i \neq 0$. The corresponding 
distance on $k[[t]]$ is complete. Indeed, denoting by $f[i]$ the term in 
$t^i$ in a series $f \in k[[t]]$, we observe that a sequence $(f_n)_{n 
\geq 0}$ is Cauchy if and only if the sequences $(f_n[i])_{n \geq 0}$ 
are all ultimately constant. A Cauchy sequence $(f_n)_{n \geq 0}$ 
therefore converges to $\sum_{i \geq 0} a_i t^i$ where $a_i$ is the 
limit of $f_n[i]$ when $n$ goes to $+\infty$.
The DVR $k[[t]]$ has a distinguished uniformizer, namely $t$. Its 
maximal ideal is then the principal ideal $(t)$ and its residue field
is canonically isomorphic to $k$. 
If one chooses $\pi = t$ and constant polynomials as representatives of 
classes modulo $t$, the expansion \eqref{eq:expandCDVR} is nothing but 
the standard writing of a formal series.
The fraction field of $k[[t]]$ is the ring of Laurent series over $k$
and, once again, the expansion \eqref{eq:expandCDVF} corresponds to the
usual writing of Laurent series.

@@

\todo{Détailler cette partie}
% parler notamment de précision absolue et de précision relative
% expliquer comment la précision se comporte vis-à-vis des opérations
% traiter un exemple de calcul de pgcd avec l'algorithme d'Euclide

@@

We recall briefly basic definitions and results about resultants and 
subresultants. For a more complete exposition, we refer to \cite[\S 
3.3]{cohen} and \cite[\S 4.1]{winkler}.
Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
$d_A$ and $d_B$ be two integers greater than or equal to the degree of 
$A$ and $B$ respectively. We consider the Sylvester application:
$$\begin{array}{rcl}
\psi : \quad
\ring_{< d_B}[X] \times \ring_{< d_A}[X] & \to &
\ring_{< d_A+d_B}[X] \smallskip \\
(U,V) & \mapsto & AU + BV
\end{array}$$
The Sylvester matrix is the matrix of $\psi$ in the canonical ordered
basis, which are
$$\begin{array}{rrl}
& ((X^{d_B-1},0), \ldots, (X,0), (1,0), (0,X^{d_A-1}), \ldots, (0,1))
& \text{for the source} \smallskip \\
\text{and} & (X^{d_A+d_B-1}, \ldots, X, 1) &
\text{for the target.}
\end{array}$$
The \emph{resultant} of $A$ and $B$ (computed in degree $d_A, d_B$) is the 
determinant of the $\psi$; we denote it by $\Res^{d_A,d_B}(A,B)$. We 
observe that it vanishes if $d_A > \deg A$ or $d_B > \deg B$. In what 
follows, we will freely drop the exponent $d_A, d_B$ if $d_A$ and $d_B$ 
are the degrees of $A$ and $B$ respectively.
Using Cramer formulae, we can build polynomials $U^{d_A, d_B}(A,B) \in
\ring_{< d_B}[X]$ and $V^{d_A, d_B}(A,B) \in \ring_{< d_A}[X]$ whose 
coefficients are, up to a sing, maximal minors of the Sylvester matrix 
such that:
$$A \cdot U^{d_A, d_B}(A,B) + B \cdot V^{d_A, d_B}(A,B) = 
\Res^{d_A, d_B}(A,B).$$
There polynomials are called the \emph{cofactors} of $A$ and $B$
(computed in degree $d_A, d_B$).

@@

Since they are defined as determinants, subresultants behave well with 
respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
and $A$ and $B$ are polynomials over $\ring$ then $\Res_j^{d_A,d_B} 
(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$ where $f(A)$ and $f(B)$ 
denotes the polynomials deduced from $A$ and $B$ respectively by 
applying $f$ coefficient-wise. This property is sometimes referred to as 
the \emph{functoriality} of subresultants.
We emphasize that, when $f$ is not injective, the relation 
$\Res_j(f(A),f(B)) = f\big(\Res_j(A,B)\big)$ does \emph{not} hold in 
general since applying $f$ may decrease the degree. Nevertheless, if
$d_A$ and $d_B$ remained fixed, this issue cannot happen.

@@

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above and assume that $d_A = \deg A$. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
$$\left\{\begin{array}{ll}
S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\varepsilon_i + 1}
s_{i-1}^{-1} \: c_i^{-\varepsilon_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 0 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\varepsilon_{i+1}} \cdot c_i^{1-\varepsilon_{i+1}}
& \text{for } i \geq -1. 
\end{array}\right.$$
Here $n_i = \deg S_i$, $\varepsilon_i = n_{i+1} - n_i$ and $s_i$ is the 
leading coefficient of $S_i$ if $i \geq 0$ and $s_{-1} = 1$ by 
convention. These sequences are finite and the above recurrence applies 
until $S_i$ has reached the value $0$.

\begin{prop}
\label{prop:prem}
With the above notations, we have:
$$\begin{array}{rcll}
\Res_j(A,B) & = & S_i & \text{if } j = n_{i-1} - 1 \smallskip \\
& = & 0 & \text{if } n_i < j < n_{i-1} - 1 < j \\
& = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\varepsilon_i-1} \cdot S_i & 
\text{if } j = n_i
\end{array}$$
for all $i$ such that $S_i$ is defined.
\end{prop}

\begin{rem}
The Proposition \ref{prop:prem} gives a formula for \emph{all} subresultants. We
note moreover that, in the common case where $n_{i-1} = n_i - 1$, the
two formulas giving $\Res_{n_i}(A,B)$ agree.

Mimicing ideas behind extended Euclide's algorithm, one can define the 
``extended subresultant pseudo-remainder sequence'' as well and obtains
recursive formulae for cofactors at the same time.
\end{rem}

Important simplifications occur in the common case where all principal 
subresultants do not vanish. Under this additional assumption, one can 
prove that the degrees of the $S_i$'s decrease by one at each step, 
meaning that we $\deg S_i = d_B - i$ for all $i$. The sequence 
$(S_i)$ then stops at $i = d_B$. Moreover, the $\varepsilon_i$'s and the 
$c_i$'s are now all ``trivial'' since we have $\varepsilon_i = 1$ and $c_i
= s_i$ for all $i$. The recurrence formula then becomes:
$$S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
\cdot (S_{i-1} \,\%\, S_i) \quad \text{for } i \geq 1.$$
and Proposition \ref{prop:prem} now simply states that $R_j = S_{d_B-j}$ 
in this case.

@@

\todo{Write a short introduction}

\todo{
Expliquer que les pertes avec l'algorithme des pseudo-restes
sont les mêmes que ceux avec l'algorithme d'Euclide.
}

@@

In this subsection, we estimate the loss of precision if we compute the 
$R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
In what follows, we are going to use a \emph{flat precision model}: this 
means that a polynomial $P(X)$ is represented as:
$$P(X) = \sum_{i=1}^n a_i X^i + O(\pi^N)
\quad \text{with } a_i \in K \text{ and } N \in \Z.$$
In other words, we assume that the software we are using does not
carry a precision data on each coefficient but only a unique precision
data for the whole polynomial. Concretely this means that, after having
computing a polynomial, the software truncates the precision on each
coefficient to the smallest precision. 
One can argue that this assumption is too strong (compared to usual
implementations of $p$-adic numbers). Nevertheless, it defines a 
simplified framework where computations can be performed and experiments 
show that it rather well reflects the behaviour of the loss of precision 
in Euclide-like algorithms.

Let $V_j$ be the valuation of the principal $j$-th subresultant of $A,B$ 
and $W_j$ be the minimum of the valuations of the coefficients of $R_j$. 
We of course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.

@@

\begin{proof}
By functoriality of subresultants, we may assume that $\ring = 
\Z[a_0, \ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$
are the two generic monic polynomials 
$A = X^d + \sum_{i=0} a_i X^i$ and
$B = X^d + \sum_{i=0} b_i X^i$.
All relations then follows from the extended subresultants 
pseudo-remainder sequence algorithm (noting in addition that all
principal subresultants are nonzero). \todo{Give details.}
\end{proof}

@@

\begin{proof} 
Given a subset $J$ of $\{0, \ldots, d-1\}$, we denote by 
$\Omega_\ring(J)$ the subset of $\Omega_\ring$ consisting of
couples $(A,B)$ for which $r_j(A,B)$ does not vanish if and only if $j 
\in J$. We want to prove that $\Omega_\ring(J)$ has cardinality 
$q^{2d-\Card J} (q-1)^{\Card J}$. To do this, we introduce several 
additional notations. First, we fix $J$ and write $J = \{n_1, \ldots, 
n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$ and set $n_{\ell+1} = 
0$ by convention. Given $n$ and $m$ two integers with $m < n$, let 
$V_{(m,n)}$ denote the set of polynomials of the form
$a_m X^m + a_{m+1} X^{m+1} \cdots + a_n X^n$
with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{(m,n)}$ has
cardinality $(q-1) q^{n-m}$. If $P$ is any polynomial of 
degree $n$ and $m < n$ is an integer, we further define $P[m{:}] \in 
V_{(m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
with coefficients $< m$. Finally, given $(A,B)$ in $\Omega_\ring$, we 
denote by $(S_i(A,B))$ its subresultant pseudo-remainder sequence as 
defined in \S \ref{subsec:subres}. We note that, if $(A,B) \in 
\Omega_\ring(J)$, the sequence $(S_i(A,B))$ stops at $i = \ell$ and we 
have $\deg S_i = n_i$ for all $i$. We now claim that the mapping
$$\begin{array}{rcl} 
\Lambda_J : \,\, 
\Omega_\ring(J) & \to & 
V_{(n_1,n_2)} \times \cdots \times V_{(n_\ell,n_{\ell+1})} \smallskip \\
(A,B) & \mapsto & 
\big(S_i(A,B)[n_{i+1}{:}]\big)_{1 \leq i \leq \ell}
\end{array}$$ 
is injective. In order to establish the claim, we remark that the 
knowledge of $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$ (for some $i$) 
is enough to reconstruct the quotient of the Euclidean division of 
$S_i(A,B)$ by $S_{i-1}(A,B)$. Thus, one can reconstruct $S_i(A,B)$ from 
the knowledge of 
$S_{i-2}(A,B)$, $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$. We deduce 
that $\Lambda_J(A,B)$ determines uniquely all $S_i(A,B)$'s 
and finally $A$ and $B$ themselves. This proves the claim.

To conclude the proof, we note that the claim implies that the cardinality of 
$\Omega_\ring(J)$ is at most $q^{2d-\ell} (q-1)^\ell$. Summing up these 
inequalities over all possible $J$, we get $\Card \Omega_\ring \leq 
q^{2d}$. This latest inequality being an equality, we need to have
$\Card \Omega_\ring(J) = q^{2d-\Card J} (q-1)^{\Card J}$ for all $J$.
\end{proof}

@@

We have seen in the previous sections that Euclide's algorithm (and 
their variants) are unstable in practice. On the other hand, 
one can compute subresultants in a very stable way by evaluating
the corresponding minors of the Sylvester matrix. Doing so, we do
not loose any significant digit. Of course, the downside is the
quite bad efficiency.

@@

In this subsection, we briefly report on the results of \cite{padicprec}. In 
\emph{loc. cit.}, the authors draw the lines of a general framework to handle a 
sharp (often optimal) track of ultrametric precision.
\todo{Give more details here.}

 In order to do so, 
they proposed to attach a single precision datum --- which is a lattice 
in a huge vector space --- to the set of all variables, instead of 
attaching a precision datum to each individual variable as we do 
usually. This proposal is supported by the following lemma in ultrametric
analysis.

@@

\todo{Add a subsection to explain the stabilization process in
general.}
